Against the Planetary Interpretation of PTFO$\,$8-8695b

### 2020-04-13 paper/ms.tex
update paper to describe data cleaning

\begin{abstract}
  PTFO$\,$8-8695b could be the youngest, shortest-period 
  hot Jupiter known.  However it has not been shown to be a planet.
  TESS recently observed PTFO$\,$8-8695 for one month.
  The TESS light-curve shows that the dominant variability in this
  system is a sinusoidal modulation with a ``long'' period $P_{\rm
  \ell}$
  of 11.96 hours, likely caused by stellar rotation.
  Also present is a complex signal, previously identified as the
  planet candidate, that repeats with a ``short'' period $P_{\rm s}$ of
  10.74 hours.
  The ``long'' and ``short'' signals show the expected beat every 4.48
  days.
  There is a dip in the complex, short-period signal.
  However ground-based photometry from the past decade shows that the
  orbital phase of the dip seems to have instantaneously jumped, at
  least once, and maybe twice.
  The TESS epoch of the dip is consistent with recent observations by
  Tanimoto et al., and differs from the discovery epoch by 5.14 hours.
  Planets do not ``jump'' in orbital phase.
  PTFO$\,$8-8695 therefore seems broadly consistent with the
  ``transient dipping'' phenomenology observed in many young M dwarfs,
  and seems unlikely to be a planet.
\end{abstract}

@@

\section{Introduction}
If it were a planet, PTFO$\,$8-8695b would be exceptional.
A transiting hot Jupiter, orbiting a $\approx$3$\,$Myr old M dwarf
in ORION would make it the youngest hot Jupiter known.
Its orbital period of only 12 HOURS would also make it the shortest
period hot Jupiter known.

@@

The 2-minute stacks for PTFO 8-8695 were then reduced to lightcurves
by the Science Processing Operations Center (SPOC) at NASA
Ames~\citep{jenkins_tess_2016}.  Our main analysis used the resulting
Presearch Data Conditioning (PDC) lightcurve, which had
non-astrophysical variability removed through the methods discussed by
\citet{smith_kepler_apertures_2017} and \citet{smith_kepler_PDC_2017}.

We also processed the 30-minute image stacks separately, as part of
the Cluster Difference Imaging
Photometric Survey (CDIPS; \citep{bouma_cluster_2019}).
The raw CDIPS light-curve shows identical features as the 2-minute
light-curve, though at 30-minute cadence.

%FIXME: these numbers
%FIXME: gotta mask orbit edges too...
Both light-curves showed a clear discontinuous ``jump'' in the last
few days, which seemed likely to be an instrumental systematic.
We masked out HOW MUCH DATA(?).
As an added precaution, we masked out the first and last N hours
of each orbit, since there is often
systematic red noise during those times.
Finally, we removed all points with non-zero quality flags \citep[see][]{tess_data_product_description_2018}.

%FIXME: these numbers
All told, we removed N\% of the original data points, and were left
with NN{,}NNN measurements of the relative flux of PTFO 8-8695.  We
normalized the data by dividing out the median flux, and then
subtracted out the mean.  Many of these and
subsequent processing steps were performed using
\texttt{astrobase}~\citep{bhatti_astrobase_2018}.

@@

Our initial inspection of the light-curve, in both its 2-minute PDCSAP
and 30-minute FFI forms, revealed an obvious beat signal
(Figure~\ref{fig:splitsignal}, top panel).

As a precursor to more detailed analysis, we calculated generalized
Lomb-Scargle periodograms using \texttt{astrobase}
\citep{lomb_1976,scargle_studies_1982,vanderplas_periodograms_2015,bhatti_astrobase_2018}.
The two largest peaks in the Lomb-Scargle periodogram of the
light-curve were clearly separated at $P_{\rm s} \approx 0.448\,{\rm
days}$ and $P_{\rm \ell} \approx 0.499\,{\rm days}$.  The $P_{\rm
\ell}$ peak had the greater power of the two.  Smaller harmonics from
each of these dominant two signals were also present.

@@

The beat peak-to-peak amplitude at maximum, when the two signals
constructively interfere, is about 14\%.  At minimum, the peak-to-peak
amplitude is about 6\%.  Assuming the signals are just two sinusoids,
algebra tells us that the peak-to-peak amplitudes should therefore be
10\% for the long-period signal, and 4\% for the short-period signal.
These numbers will turn out to be over-estimates, but only slightly.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.98\textwidth]{f1.png}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf TESS light-curve of PTFO 8-8695 (Sector 6, Orbit 1).}
	{\it Top}: ``Raw'' \texttt{PDCSAP} mean-subtracted relative flux versus time. The beat period of 4.48 days is visible by eye.
	The model plotted underneath the data includes 2 harmonics at the long
	period $P_{\rm \ell}$, plus 2 harmonics and a transit at the short period $P_{\rm s}$.
	{\it Upper middle}: Long-period signal, equal to the raw signal minus the short-period signal.
	{\it Lower middle}: Short-period signal, equal to the raw signal minus the long-period signal.
	{\it Bottom}: residual.
	The data are binned from 2 to 10 minute cadence as a convenience for plotting and fitting.
		\label{fig:splitsignal}
	}
\end{figure*}

@@

We opted to model the light-curve as a linear combination of Fourier
harmonics at the short and long periods, plus a transit at the short
period.  Symbolically,
\begin{equation}
  f = f_{\rm s} + f_{\rm \ell}
  = f_{\rm transit,s} + f_{\rm Fourier,s} + f_{\rm Fourier,\ell},
\end{equation}
where $f_{\rm s}$ is the relative flux at the short period, and
$f_{\rm \ell}$ is the flux at the long period.  Writing out the
Fourier terms,
\begin{align}
  f = &f_{\rm transit,s} + \sum_{n=1}^{n=N} A_n \sin(n\omega_{\rm s}t)
  + \sum_{n=1}^{n=N} B_n \cos(n\omega_{\rm s}t)\\
  &+ \sum_{m=1}^{m=M} A_m \sin(m[\omega_{\rm \ell}t+\phi_{\rm \ell}])
  + \sum_{m=1}^{m=M} B_m \cos([m\omega_{\rm \ell}t+\phi_{\rm \ell}]), \nonumber
\end{align}
for $N$ and $M$ the total number of harmonics at the short and long
period, respective, $A_i$ and $B_i$ for each harmonic term, and
$\omega_i = 2\pi / P_i$ the angular frequency for $i$ the short or
long period index.  We fixed the ``phase-offset'' for the short period
signal to be zero, and let the reference time for the long period
signal float by introducing $\phi_{\rm \ell}$.  Since we did not a
priori know how many harmonics would be appropriate, we considered a
number of different choices for $N$ and $M$, and used the Bayesian
information criterion to choose the appropriate model.

@@

The TESS pixels are $\approx21$'' per side, and so we need to consider
whether light from neighboring stars could affect the photometry.  The
scene is shown in Figure~\ref{fig:scene}.  
The pixels used to
measure the background level are indicated with an `\texttt{X}' hatch,
and the pixels used for the final light-curve are shown with the
`\texttt{/}' hatch.

@@

First, we examined the CDIPS full frame image light-curves of the
target, which are available on MAST \citep{bouma_cluster_2019}.
The maximal peak-to-peak beat amplitude is consistently $\approx$10\%
across apertures of radii 1, 1.5, and 2.25 pixels.
If Star A were the source of the long-period variability, we would expect the
peak variability amplitude to be smallest in the 1 pixel aperture, based on the
separation of the sources (Figure~\ref{fig:scene}, bottom).
From this test alone, it seems unlikely that Star A is the source of
the long-period signal.

Second, we examined the light-curve of each pixel in the scene
individually.  We opted to use the
interactive tools implemented in
\texttt{lightkurve} \citep{lightkurve_2018}.  If Star A were the
source of the long-period variability, we would expect the pixels
nearest to Star A to show a sinusoidal signal with
amplitude exceeding $10\%$.  We find no evidence for
this being the case.  The pixel directly below Star A does not
clearly show the sinusoidal variability, and the peak-to-peak 
variability in that pixel is $\lesssim 8\%$.  In contrast, the
south-easternmost pixel within PTFO 8-8695's aperture (the pixel 
furthest from Star A that was used in the optimal aperture) shows the $P_{\rm \ell}$ sinusoidal
variability signal at $\approx 10\%$ amplitude.

>>>>

\begin{abstract}
  PTFO$\,$8-8695b could be the youngest, shortest-period 
  hot Jupiter known.  However it has not been shown to be a planet.
  TESS recently observed PTFO$\,$8-8695 for one month.
  The TESS lightcurve shows that the dominant variability in this
  system is a sinusoidal modulation with a ``long'' period $P_{\rm
  \ell}$
  of 11.96 hours, likely caused by stellar rotation.
  Also present is a complex signal, previously identified as the
  planet candidate, that repeats with a ``short'' period $P_{\rm s}$ of
  10.74 hours.
  The ``long'' and ``short'' signals show the expected beat every 4.48
  days.
  There is a dip in the complex, short-period signal.
  However ground-based photometry from the past decade shows that the
  orbital phase of the dip seems to have instantaneously jumped, at
  least once, and maybe twice.
  The TESS epoch of the dip is consistent with recent observations by
  Tanimoto et al., and differs from the discovery epoch by 5.14 hours.
  Planets do not ``jump'' in orbital phase.
  Given the available evidence,
  PTFO$\,$8-8695 seems consistent with the
  ``transient dipping'' phenomenology observed in many young M dwarfs.
  It seems rather unlikely to be a planet.
\end{abstract}

@@

\section{Introduction}
If PTFO$\,$8-8695b were a planet, it would be exceptional.
A transiting hot Jupiter, orbiting a $\approx$3$\,$Myr old M dwarf
in ORION would make it the youngest hot Jupiter known.
Its orbital period of only 12 HOURS would also make it the shortest
period hot Jupiter known.

@@

The 2-minute stacks for PTFO 8-8695 were then reduced to lightcurves
by the Science Processing Operations Center (SPOC) at NASA
Ames~\citep{jenkins_tess_2016}.  Our main analysis used the resulting
Presearch Data Conditioning (PDC) lightcurve.  The PDC lightcurve
aperture used pixels chosen to maximize the SNR of the total flux of
the target \citep{smith_kepler_apertures_2017}.  Non-astrophysical
variability was removed through the methods discussed by
\citet{smith_kepler_PDC_2017}.

As an independent check on the shorter cadence SPOC light-curve, we
separately processed the 30-minute image stacks as part of the Cluster
Difference Imaging Photometric Survey (CDIPS;
\citep{bouma_cluster_2019}).  The CDIPS lightcurve used a circular
aperture with radius 1 pixel.

For data cleaning, we removed all points with non-zero quality flags
\citep[{\it e.g.},][]{tess_data_product_description_2018}.  We also
masked out the first and last 6 hours of each orbit, since there is
often systematic red noise during those times.  Both the CDIPS and PDC
lightcurves showed a clear discontinuous ``jump'' in the last few days
of orbit 20, which seemed likely to be an instrumental systematic.  We
correspondingly masked out times from BJD 2458488.3 until the end of
the orbit.  The PDC lightcurve initially had 15{,}678 points.  The
quality cut removed 854, masking the orbit edges removed 716, and
removing the final few days of orbit 20 removed 1079.  83\% of the
highest-quality data remained.

We normalized these points by dividing out the median flux. We then
subtracted by unity to simplify subsequent analysis.  Many of these
and subsequent processing steps were performed using
\texttt{astrobase}~\citep{bhatti_astrobase_2018}.

@@

Our initial inspection of the lightcurve, in both its 2-minute PDCSAP
and 30-minute FFI forms, revealed an obvious beat signal
(Figure~\ref{fig:splitsignal}, top panel).

As a precursor to more detailed analysis, we calculated generalized
Lomb-Scargle periodograms using \texttt{astrobase}
\citep{lomb_1976,scargle_studies_1982,vanderplas_periodograms_2015,bhatti_astrobase_2018}.
The two largest peaks in the Lomb-Scargle periodogram of the
lightcurve were clearly separated at $P_{\rm s} \approx 0.448\,{\rm
days}$ and $P_{\rm \ell} \approx 0.499\,{\rm days}$.  The $P_{\rm
\ell}$ peak had the greater power of the two.  Smaller harmonics from
each of these dominant two signals were also present.

@@

The beat peak-to-peak amplitude at maximum, when the two signals
constructively interfere, is about 14\%.  At minimum, the peak-to-peak
amplitude is about 6\%.  Assuming the signals are just two sinusoids,
algebra tells us that the peak-to-peak amplitudes should therefore be
10\% for the long-period signal, and 4\% for the short-period signal.
These order-of-magnitude numbers will turn out to be roughly correct.

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.98\textwidth]{f1.png}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf TESS lightcurve of PTFO 8-8695 (Sector 6, Orbit 1).}
	{\it Top}: ``Raw'' \texttt{PDCSAP} mean-subtracted relative flux versus time. The beat period of 4.48 days is visible by eye.
	The model plotted underneath the data includes 2 harmonics at the long
	period $P_{\rm \ell}$, plus 2 harmonics and a transit at the short period $P_{\rm s}$.
	{\it Upper middle}: Long-period signal, equal to the raw signal minus the short-period signal.
	{\it Lower middle}: Short-period signal, equal to the raw signal minus the long-period signal.
	{\it Bottom}: residual.
	The data are binned from 2 to 10 minute cadence as a convenience for plotting and fitting.
		\label{fig:splitsignal}
	}
\end{figure*}

@@

We opted to model the lightcurve as a linear combination of Fourier
harmonics at the short and long periods, plus a transit at the short
period.  Symbolically,
\begin{equation}
  f = f_{\rm s} + f_{\rm \ell}
  = f_{\rm transit,s} + f_{\rm Fourier,s} + f_{\rm Fourier,\ell},
\end{equation}
where $f_{\rm s}$ is the relative flux at the short period, and
$f_{\rm \ell}$ is the flux at the long period.  Writing out the
Fourier terms,
\begin{align}
  f = &f_{\rm transit,s} + \sum_{n=1}^{n=N} A_n \sin(n\omega_{\rm s}t)
  + \sum_{n=1}^{n=N} B_n \cos(n\omega_{\rm s}t)\\
  &+ \sum_{m=1}^{m=M} A_m \sin(m[\omega_{\rm \ell}t+\phi_{\rm \ell}])
  + \sum_{m=1}^{m=M} B_m \cos(m[\omega_{\rm \ell}t+\phi_{\rm \ell}]), \nonumber
\end{align}
for $N$ and $M$ the total number of harmonics at the short and long
period, respective, $A_i$ and $B_i$ for each harmonic term, and
$\omega_i = 2\pi / P_i$ the angular frequency for $i$ the short or
long period index.  We fixed the ``phase-offset'' for the short period
signal to be zero, and let the reference time for the long period
signal float by introducing $\phi_{\rm \ell}$.  Since we did not a
priori know how many harmonics would be appropriate, we considered a
number of different choices for $N$ and $M$, and used the Bayesian
information criterion to choose the appropriate model.

@@

The TESS pixels are $\approx21$'' per side, and so we need to consider
whether light from neighboring stars could affect the photometry.  The
scene is shown in Figure~\ref{fig:scene}.  
The pixels used to
measure the background level are indicated with an `\texttt{X}' hatch,
and the pixels used for the final lightcurve are shown with the
`\texttt{/}' hatch.

@@

First, we examined the CDIPS full frame image lightcurves of the
target, which are available on MAST \citep{bouma_cluster_2019}.
The maximal peak-to-peak beat amplitude is consistently $\approx$10\%
across apertures of radii 1, 1.5, and 2.25 pixels.
If Star A were the source of the long-period variability, we would expect the
peak variability amplitude to be smallest in the 1 pixel aperture, based on the
separation of the sources (Figure~\ref{fig:scene}, bottom).
From this test alone, it seems unlikely that Star A is the source of
the long-period signal.

Second, we examined the lightcurve of each pixel in the scene
individually.  We opted to use the
interactive tools implemented in
\texttt{lightkurve} \citep{lightkurve_2018}.  If Star A were the
source of the long-period variability, we would expect the pixels
nearest to Star A to show a sinusoidal signal with
amplitude exceeding $10\%$.  We find no evidence for
this being the case.  The pixel directly below Star A does not
clearly show the sinusoidal variability, and the peak-to-peak 
variability in that pixel is $\lesssim 8\%$.  In contrast, the
south-easternmost pixel within PTFO 8-8695's aperture (the pixel 
furthest from Star A that was used in the optimal aperture) shows the $P_{\rm \ell}$ sinusoidal
variability signal at $\approx 10\%$ amplitude.
